---
title: "How visualizing inferential uncertainty can mislead readers about treatment effects in scientific results: Figures and analyses for CHI 2020 paper"
author: "Jake M. Hofman, Daniel G. Goldstein, Jessica Hullman"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 3
    fig_width: 6
    fig_height: 5
---


```{r setup, include = F}
library(gridExtra)
library(tidyverse)
library(scales)
library(broom)
library(lsr)
library(knitr)
library(lsmeans)


subplot_results <- function(plot_data, num_se = 1, show_facets = T, y_label = "Outcome", use_sd = F, dodge_width = 0.2, coord_flip = F, true_effects = NA) {
  bar_type <- unique(plot_data$bar_type)
  interval_size <- unique(plot_data$interval_size)
  
  # set symbol and color
  if (bar_type == "pi") {
    color <- "grey 50"
    shape <- 17
  } else if (bar_type == "ci") {
    color <- "black"
    shape <- 16
  } else {
    stop("shouldn't happen")
  }
      
  p <- ggplot(plot_data, aes(y = mean_outcome, x = condition))
  
  if (is.data.frame(true_effects)) {
    p <- p + geom_hline(data = true_effects, 
                        aes(yintercept = auc), linetype = 'dashed', alpha = 0.25)
  }
  
  p <- p +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0, color = color) +
    geom_point(position = position_dodge(width = dodge_width), size = 2, color = color, shape = shape)
  
  # set title
  if (num_se == 1) {
    if (bar_type == "pi")
      title <- "The same results with error bars showing one standard deviation"
    else if (bar_type == "ci")
      title <- "Results with error bars showing one standard error"
    else
      stop("shouldn't happen")
  } else {
    if (bar_type == "pi")
      title <- sprintf("The same results with error bars showing show %d%% of responses", round(100*interval_size))
    else if (bar_type == "ci")
      title <- sprintf("Results with error bars showing %d%% CI on the mean", round(100*interval_size))
  }
  
  p <- p + labs(subtitle = title, x = '', y = y_label) +
    theme(legend.title = element_blank(),
          legend.position = "none",
          plot.subtitle = element_text(size=9))
  
  if (coord_flip == T) {
    p <- p + coord_flip()
  } else {
    p <- p + theme(axis.title.x = element_blank())
  }
  
  if (show_facets == T)
    p <- p + facet_wrap(~ facet_var)
  
  p
}

plot_results <- function(df, num_se = 1, show_facets = T, y_label = "Outcome", use_sd = F, dodge_width = 0.2, coord_flip = F, true_effects = NA) {
  # determine interval size and upper and lower percentiles
  interval_size <- pnorm(num_se) - pnorm(-num_se)
  lower_percentile <- (1 - interval_size) / 2
  upper_percentile <- 1 - lower_percentile
  
  # compute summary stats
  summary_stats <- df %>%
    group_by(facet_var, condition) %>%
    summarize(mean_outcome = mean(outcome),
              sd_outcome = sd(outcome),
              se_outcome = sd(outcome) / sqrt(n()),
              pi_lower = ifelse(use_sd == T, mean_outcome - sd_outcome,
                                quantile(outcome, lower_percentile)),
              pi_upper = ifelse(use_sd == T, mean_outcome + sd_outcome,
                                quantile(outcome, upper_percentile)),
              ci_lower = mean_outcome - num_se*se_outcome,
              ci_upper = mean_outcome + num_se*se_outcome) %>%
    ungroup()
  
  # separate out confidence interval
  plot_data_1 <- summary_stats %>%
    select(condition, facet_var, mean_outcome, lower = ci_lower, upper = ci_upper) %>%
    mutate(bar_type = "ci", interval_size = interval_size)
  
  # separate out prediction interval
  plot_data_2 <- summary_stats %>%
    select(condition, facet_var, mean_outcome, lower = pi_lower, upper = pi_upper) %>%
    mutate(bar_type = "pi", interval_size = interval_size)
  
  p1 <- subplot_results(plot_data_1, num_se, show_facets, y_label, use_sd, dodge_width, coord_flip, true_effects)
  p2 <- subplot_results(plot_data_2, num_se, show_facets, y_label, use_sd, dodge_width, coord_flip, true_effects)
  
  grid.arrange(p1, p2, nrow = 2)
}

```

# Set constants
```{r set-constants}
# set style
theme_set(theme_bw() +
            theme(panel.grid.major=element_blank(),
                  panel.grid.minor=element_blank()))

# set figure directory
figure_dir <- "figures"

# set flag for whether to use +/- 1sd next to 1se bars
# if true, bars are mean +/- 1sd
# if false, bars span 68% of responses
use_1sd_for_1se <- T
```

# Experiment 1

## Load data

```{r exp1-load-data}
load('data/experiment_1_clean.Rdata')
exp1 <- df
rm(df)

exp1 <- exp1 %>%
  mutate(condition = recode_factor(as.factor(condition),
                                   SE = "Visualization\nshowed 95% CI",
                                   SD = "Visualization\nshowed 95% PI"),
         condition = factor(as.character(condition), levels = rev(levels(condition))),
         text_condition = recode_factor(as.factor(text_condition),
                                        show_both_stats = "Text contained extra information",
                                        show_viz_stats_only = "Text matched visualization"),
         text_condition = factor(as.character(text_condition), levels = rev(levels(text_condition))))

exp1_true_dists <- true_dists
exp1_true_effects <- true_effects
rm(true_dists, true_effects)
```

## Remove participants who failed comprehension test

```{r exp1-filter}
nrow(exp1)
exp1 <- exp1 %>% filter(superiority_standard == 0.5)
nrow(exp1)
```

## Willingness to pay

### T-test
```{r exp1-wtp-t-test}
exp1 %>%
  group_by(text_condition) %>%
  do(tidy(t.test(wtp_final ~ condition, data = .))) %>%
  kable()
```

### Effect size
```{r exp1-wtp-effect-size}
exp1 %>%
  group_by(text_condition) %>%
  do(data.frame(cohens_d = cohensD(formula = wtp_final ~ condition, data = .))) %>%
  kable()
```

### Plots with 1 SE / 68% intervals

```{r exp1-wtp-1se}
p <- exp1 %>%
  rename(outcome = wtp_final, facet_var = text_condition) %>%
  plot_results(num_se = 1, show_facets = T, y_label = "Willingness to pay", use_sd = use_1sd_for_1se, coord_flip = T)
ggsave(sprintf('%s/exp1-wtp-1se.pdf', figure_dir), p, width = 5, height = 4)
```

## Probability of superiority

### T-test
```{r exp1-special-t-test}
exp1 %>%
  group_by(text_condition) %>%
  do(tidy(t.test(superiority_special ~ condition, data = .))) %>%
  kable()
```

### Effect size
```{r exp1-special-effect-size}
exp1 %>%
  group_by(text_condition) %>%
  do(data.frame(cohens_d = cohensD(formula = superiority_special ~ condition, data = .))) %>%
  kable()
```

### Plots with 1 SE / 68% intervals

```{r exp1-prob-superiority-1se}
p <- exp1 %>%
  rename(outcome = superiority_special, facet_var = text_condition) %>%
  plot_results(num_se = 1, show_facets = T, y_label = "Probability of superiority", use_sd = use_1sd_for_1se, true_effects = exp1_true_effects, coord_flip = T)
ggsave(sprintf('%s/exp1-prob-superiority-1se.pdf', figure_dir), p, width = 5, height = 4)
```

# Experiment 2

## Load data

```{r exp2-load-data}
load('data/experiment_2_clean.Rdata')
exp2 <- df
rm(df)

exp2 <- exp2 %>%
  mutate(condition = recode_factor(condition,
                                   HOPS = 'Visualization showed\nanimated samples (HOPs)',
                                   SD = "Visualization showed\n95% PI",
                                   SE_rescaled = "Visualization showed\n95% CI rescaled",
                                   SE = "Visualization showed\n95% CI"))

true_dists <- true_dists %>%
  mutate(distribution = recode_factor(distribution,
                                      distribution_special = "Special boulder",
                                      distribution_standard = "Standard boulder"))  

distbuilder_values_by_worker <- distbuilder_values_by_worker %>%
  ungroup() %>%
  mutate(condition = recode_factor(condition,
                                   HOPS = 'Visualization showed\nanimated samples (HOPs)',
                                   SD = "Visualization showed\n95% PI",
                                   SE_rescaled = "Visualization showed\n95% CI rescaled",
                                   SE = "Visualization showed\n95% CI"),
         distribution = recode_factor(distribution,
                                      distribution_special = "Special boulder",
                                      distribution_standard = "Standard boulder"))

exp2_true_dists <- true_dists
exp2_true_effects <- true_effects %>%
  mutate(facet_var = effect_size)
rm(true_dists, true_effects)
```

## Remove participants who failed comprehension test

```{r exp2-filter}
nrow(exp2)

exp2 <- exp2 %>%
  filter(superiority_standard == .5)

distbuilder_values_by_worker <- distbuilder_values_by_worker %>%
  filter(superiority_standard == .5)

nrow(exp2)
```

## Willingness to pay

### Pairwise T-tests
```{r exp2-wtp-t-test}
exp2 %>%
  group_by(effect_size) %>%
  do(tidy(pairwise.t.test(.$wtp_final, .$condition, data = .))) %>%
  mutate(group1 = gsub('\n',' ', group1), group2 = gsub('\n',' ', group2)) %>%
  kable()
```

### Anova 

```{r exp2-wtp-anova}
model <- aov(wtp_final ~ effect_size + condition, data = exp2)
summary(model)
pairs(lsmeans(model, "condition"), adjust = "none")
```

### Plots with 1 SE / 68% intervals

```{r exp2-wtp-1se}
p <- exp2 %>%
  rename(outcome = wtp_final, facet_var = effect_size) %>%
  plot_results(num_se = 1, show_facets = T, y_label = "Willingness to pay", use_sd = use_1sd_for_1se, coord_flip = T, dodge_width = 0.5)
ggsave(sprintf('%s/exp2-wtp-1se.pdf', figure_dir), p, width = 6, height = 5)
```


## Probability of superiority

### Pairwise T-tests

```{r exp2-special-t-test}
exp2 %>%
  group_by(effect_size) %>%
  do(tidy(pairwise.t.test(.$superiority_special, .$condition, data = .))) %>%
  mutate(group1 = gsub('\n',' ', group1), group2 = gsub('\n',' ', group2)) %>%
  kable()
```

### Anova 

```{r exp2-superiority-special-anova}
model <- aov(superiority_special ~ effect_size + condition, data = exp2)
summary(model)
pairs(lsmeans(model, "condition"), adjust = "none")
```


### Plots with 1 SE / 68% intervals

```{r exp2-prob-superiority-1se}
p <- exp2 %>%
  rename(outcome = superiority_special, facet_var = effect_size) %>%
  plot_results(num_se = 1, show_facets = T, y_label = "Stated probability of superiority", use_sd = use_1sd_for_1se, true_effects = exp2_true_effects, dodge_width = 0.5, coord_flip = T)
ggsave(sprintf('%s/exp2-prob-superiority-1se.pdf', figure_dir), p, width = 6, height = 5)
```


## Aggregate distbuilder histogram

```{r exp2-distbuilder-histogram, fig.width=9, fig.height=3}

bin_size <- mean(diff(bin_labels$label))

plot_data <- merge(bin_labels, exp2_true_dists, all.x = T, all.y = T) %>%
  mutate(lower = label,
         center = label + bin_size / 2,
         upper = label + bin_size,
         true_frac = pnorm(upper, mean = mean, sd = sd) - pnorm(lower, mean = mean, sd = sd))

distbuilder_values_by_worker %>%
  group_by(distribution, effect_size, condition, bin) %>%
  summarize(total_count = sum(count)) %>%
  group_by(distribution, effect_size, condition) %>%
  mutate(bin_center = bin + bin_size / 2,
         frac = total_count / sum(total_count)) %>%
  ggplot(aes(x = bin_center, y = frac, fill = distribution)) +
    geom_line(data = plot_data, aes(x = center, y = true_frac, color = distribution), linetype = 'solid') +
    geom_vline(data = exp2_true_dists,
               aes(xintercept = mean, color = distribution),
               linetype = 'dotted') +
    geom_bar(stat = "identity", position = "identity", alpha = 0.5) +
    facet_grid(effect_size ~ condition) +
    scale_fill_discrete(breaks = c('Standard boulder', 'Special boulder')) +
    scale_color_discrete(breaks = c('Standard boulder', 'Special boulder')) +
    labs(x = 'Distance',
         y = 'Fraction of responses') +
    theme(legend.title = element_blank())
ggsave(sprintf('%s/exp2-distbuilder-histogram.pdf', figure_dir), width = 9, height = 3)
```

## Implied standard deviations from distribution builder
```{r distbuilder_extract_values}

distbuilder_summary_stats_by_worker <- distbuilder_values_by_worker %>%
  group_by(worker_id, condition, effect_size, distribution) %>%
  summarize(mean = sum(count * bin) / sum(count),
            sd = sqrt(sum(count * (bin - mean)^2) / (sum(count - 1)))) %>%
  ungroup() %>%
  mutate(condition_bin = ifelse(grepl('^SE', condition),
                                'Sampling distribution\n(SE, SE rescaled)',
                                'Population distribution\n(SD, HOPS)'))

```

### Anova 

```{r implied_sd_anova}
model <- aov(sd ~ effect_size + condition, data = distbuilder_summary_stats_by_worker)
summary(model)
pairs(lsmeans(model, "condition"), adjust = "none")
```

## Implied probability of superiority

### Compute AUC

```{r exp2-distbuilder-compute-auc, warning=FALSE, message=FALSE}
implied_auc_by_worker <- distbuilder_values_by_worker %>%
  group_by(worker_id, condition, effect_size) %>%
  do({
    data.frame(distribution = rep(.$distribution, .$count),
               x = rep(.$bin, .$count))
  }) %>%
  yardstick::roc_auc(distribution, x) %>%
  select(-.metric, -.estimator, auc = .estimate) %>%
  mutate(condition_bin = ifelse(grepl('^SE', condition),
                                'Sampling distribution\n(SE, SE rescaled)',
                                'Population distribution\n(SD, HOPS)')) %>%
  ungroup()
```

### Anova

```{r implied_prob_superiority_anova}
model <- aov(auc ~ effect_size + condition, data = implied_auc_by_worker)
summary(model)
pairs(lsmeans(model, "condition"), adjust = "none")
```

### Plots with 1 SE / 68% intervals
```{r exp2-distbuilder-prob-superiority-1se}
p <- implied_auc_by_worker %>%
  rename(outcome = auc, facet_var = effect_size) %>%
  plot_results(num_se = 1, show_facets = T, y_label = "Implied probability of superiority", use_sd = use_1sd_for_1se, true_effects = exp2_true_effects, dodge_width = 0.5, coord_flip = T)
ggsave(sprintf('%s/exp2-distbuilder-prob-superiority-1se.pdf', figure_dir), p, width = 6, height = 5)
```


# R Session information with package versions

```{r session-info}

sessionInfo()

```